---
title: "Statistical assignment 4"
author: "James Hughes, 660056969"
date: "27/02/2020"
output: github_document
---

```{r setup, include=FALSE}
# Please note these options.
# This tells R Markdown that we want to show code in the output document.
knitr::opts_chunk$set(echo = TRUE)
# Switching off messages in the output document. 
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
# Switching on caching to make things faster (don't commit cache files on Github).
knitr::opts_chunk$set(cache = TRUE)
```

In this assignment you will need to reproduce 5 ggplot graphs. I supply graphs as images; you need to write the ggplot2 code to reproduce them and knit and submit a Markdown document with the reproduced graphs (as well as your .Rmd file).

First we will need to open and recode the data. I supply the code for this; you only need to change the file paths.

```{r}

library(tidyverse)
Data8 <- read_tsv("C:/Users/James/Documents/Data III/assignment-4-jameshughes96/UKDA-6614-tab/tab/ukhls_w8/h_indresp.tab")
Data8 <- Data8 %>%
        select(pidp, h_age_dv, h_payn_dv, h_gor_dv)
Stable <- read_tsv("C:/Users/James/Documents/Data III/assignment-4-jameshughes96/UKDA-6614-tab/tab/ukhls_wx/xwavedat.tab")
Stable <- Stable %>%
        select(pidp, sex_dv, ukborn, plbornc)
Data <- Data8 %>% left_join(Stable, "pidp")
rm(Data8, Stable)
Data <- Data %>%
        mutate(sex_dv = ifelse(sex_dv == 1, "male",
                           ifelse(sex_dv == 2, "female", NA))) %>%
        mutate(h_payn_dv = ifelse(h_payn_dv < 0, NA, h_payn_dv)) %>%
        mutate(h_gor_dv = recode(h_gor_dv,
                         `-9` = NA_character_,
                         `1` = "North East",
                         `2` = "North West",
                         `3` = "Yorkshire",
                         `4` = "East Midlands",
                         `5` = "West Midlands",
                         `6` = "East of England",
                         `7` = "London",
                         `8` = "South East",
                         `9` = "South West",
                         `10` = "Wales",
                         `11` = "Scotland",
                         `12` = "Northern Ireland")) %>%
        mutate(placeBorn = case_when(
                ukborn  == -9 ~ NA_character_,
                ukborn < 5 ~ "UK",
                plbornc == 5 ~ "Ireland",
                plbornc == 18 ~ "India",
                plbornc == 19 ~ "Pakistan",
                plbornc == 20 ~ "Bangladesh",
                plbornc == 10 ~ "Poland",
                plbornc == 27 ~ "Jamaica",
                plbornc == 24 ~ "Nigeria",
                TRUE ~ "other")
        )

```

Reproduce the following graphs as close as you can. For each graph, write two sentences (not more!) describing its main message.  

1. Univariate distribution (20 points).

```{r}

ggplot(data=Data, aes(x=h_payn_dv)) +
    geom_freqpoly() +
    xlab("Net monthly pay") + 
    ylab("Number of respondents")
    
```
This graph shows the distribution of respondents based on their net monthly pay, with the majority of respondents earning a net monthly pay betwen £0 and £4000 a month. The largest proportion of earners take home between £1000 and £1500 pounds a month, with very few respondents earning more than £4000 a month. 

2. Line chart (20 points). The lines show the non-parametric association between age and monthly earnings for men and women.

```{r}

Data$Sex <- Data$sex_dv

ggplot(Data, aes(x = h_age_dv, y = h_payn_dv, color = Sex)) +
    geom_smooth(aes(linetype = Sex)) +
    scale_linetype_manual(values=c("solid", "dashed")) +
    scale_color_manual(values = c("black", "black")) +
    xlim(16, 65) +
    xlab("Age") +
    ylab("Monthly earnings")

```
Between the ages of 16 and 44 for males and the ages of 16 and 48 for females, there is a positive correlation between age and monthly earnings, after which monthly earnings begin to decrease with age. Men and women earn similar amounts between the ages of 16 and 25, after which males earn significantly more than women from thereon in. 

3. Faceted bar chart (20 points).

```{r}
Data2 <- Data %>%
    group_by(sex_dv, placeBorn) %>% 
           summarise(medianIncome = median(h_payn_dv, na.rm = TRUE)) %>% 
    filter(!is.na(sex_dv), !is.na(placeBorn))

ggplot(Data2, aes(x=sex_dv, y=medianIncome)) +
    geom_bar(stat = "Identity") +
    facet_wrap(~ placeBorn) +
    xlab("Sex") +
    ylab("Median monthly net pay")

```
This graph shows that regardless of where an individual is born, the median monthly net income is less for females than it is for males. The graph also shows that median monthly net income varies depending on an individual's birthplace, with those born in Bangladesh having the lowest median monthly income, and those born in Ireland having the highest median monthly income. 


4. Heat map (20 points).

```{r}

heatmap <- Data %>%
    group_by(h_gor_dv, placeBorn) %>% 
    summarise(
        MeanAge = mean(h_age_dv)
    ) %>% 
    filter(!is.na(h_gor_dv), !is.na(placeBorn))

ggplot(heatmap, aes(x = h_gor_dv, y = placeBorn, fill= MeanAge)) + 
  geom_tile() +
    xlab("Region") +
    ylab("Country of birth") +
    labs(fill = "Mean Age") +
    theme(axis.text.x=element_text(angle=90, hjust=1)) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "white"))

```
This heat map identifies that the youngest individuals in the study were those that live in Scotland but were born in Nigeria, with a mean age of around 30, and in general, individuals from Jamaica and Ireland were older than most respondents. The oldest individuals in the study were those born in Jamaica but living in Scotland, with a mean age of over 80 (although these extremes are likely due to a low number of respondents meeting the criteria), and there were certain respondents that were missing from the study entirely, represented by the blank squares (such as those born in Bangladesh living in Northern Ireland). 


5. Population pyramid (20 points).

```{r}

ggplot(data=Data,aes(x=h_age_dv, fill=sex_dv)) + 
  geom_bar(data=subset(Data, Data$sex_dv=="female")) + 
  geom_bar(data=subset(Data, Data$sex_dv=="male"), aes(y=..count..*(-1))) +
    scale_fill_manual(values = c("red3", "steelblue")) +
  scale_y_continuous() +
    coord_flip() +
    xlab("Age") +
    ylab("n") +
    labs(fill = "Sex") +
    theme_bw()

```
This population pyramid shows the distribution of respondents based on their age and gender, and from looking at the graph it appears that there are more female respondents in the study than males across the entire age distribution. The largest concentration of ages in the study for both genders is between the ages of 45 and 55, with a lack of respondents aged between 25 and 30, and the number of respondents tapering down above the age of 75. 












